---
output: 
  pdf_document: 
    toc: yes
    toc_depth: 5
    fig_caption: true
    keep_tex: true
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
mainfont: Times New Roman
header-includes:
  - \usepackage{fancyvrb}
  - \fvset{fontsize=\tiny}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{geometry}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{Verbatim}{Verbatim}{fontsize=\footnotesize}
  - \usepackage[hyphens]{url}
  - \usepackage{microtype} 
csl: apa.csl
---
\newpage 

With the previous steps of a DE analysis now we can narrow down the formal DE analysis for this we are using the [limma](https://bioconductor.org/packages//release/bioc/html/limma.html) package, this package adjust linear models to the genes to acces linear regretions about its expression and then adjust that values to model the diferential expression genes, becaouse it accounts with large amounts of testing it retruves a padj value associated with the log2 fold changues. 

```{r}
#chargue the neccesary libraries
library(limma)
library(SummarizedExperiment)
library(ggplot2)

#cragamos los obejtos necesarios
load("Results/objects_R/rse_gene_SRP119675_preprocessed.Rdata")
load("Results/objects_R/rse_gene_SRP119675_dge_normalized.Rdata")
```

### Defining a model 
As we could see in the preprocessing part of the analysis the variables that are relavant for explaining variance in some genes that would be of interest are: 
- sra_attribute.genotype --> as the principal 
- sra_attribute.age 
- sra_attribute.gender 

So with this in mind we can define our `model.matrix` object for account for our estadistical assuptions 

```{r}
#define the object
mod <- model.matrix(
  ~ sra_attribute.genotype + sra_attribute.age + sra_attribute.gender,
  data = colData(rse_gene_SRP119675_filtered)
)

#visualize the attributes we are modeling
print(colnames(mod))

#we can compare this with our levels of our variables to know which is the reference used for the intercept, because for default analysis modelmatrix use the first level of the attribute:
print("The reference levels or baseline we are using are")

cat(
  "For the genotype:\n",
  levels(colData(rse_gene_SRP119675_filtered)$sra_attribute.genotype)[1],
  "\n"
)

cat(
  "For the age:\n",
  levels(colData(rse_gene_SRP119675_filtered)$sra_attribute.age)[1],
  "\n"
)

cat(
  "For the gender:\n",
  levels(colData(rse_gene_SRP119675_filtered)$sra_attribute.gender)[1],
  "\n"
)
``` 

We can see that the baseline for our firt variable genotype is the 5xFAD, beacuse it is a genotype that is used for exploring alzheimer disease in mouse (3 APP mutations) it shoul be more convinient to changue it for the wildtype 

```{r}
#redifine the first element of the levels
colData(rse_gene_SRP119675_filtered)$sra_attribute.genotype <- relevel(
  colData(rse_gene_SRP119675_filtered)$sra_attribute.genotype,
  ref = "WT"
)

cat(
  "The new baselinen for the genotype would be: \n",
  levels(colData(rse_gene_SRP119675_filtered)$sra_attribute.genotype)[1],
  "\n"
)

#redifine de model
mod <- model.matrix(
  ~ sra_attribute.genotype + sra_attribute.age + sra_attribute.gender,
  data = colData(rse_gene_SRP119675_filtered)
)

#visualize the attributes we are modeling
print(colnames(mod))

```

Now we can fit this model to the linear one that use limma 

```{r}
vGene <- voom(dge, mod, plot = TRUE)

eb_results <- eBayes(lmFit(vGene))

de_results <- topTable(
  eb_results,
  coef = 2,
  number = nrow(rse_gene_SRP119675_filtered),
  sort.by = "none"
)
dim(de_results)
``` 

#### Someplots 

```{r}
plotMA(eb_results, coef = 2)

volcanoplot(eb_results, coef = 4, highlight = 5, names = de_results$gene_name)
```