---
output: 
  pdf_document: 
    toc: yes
    toc_depth: 5
    fig_caption: true
    keep_tex: true
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
mainfont: Times New Roman
header-includes:
  - \usepackage{fancyvrb}
  - \fvset{fontsize=\tiny}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{geometry}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{Verbatim}{Verbatim}{fontsize=\footnotesize}
  - \usepackage[hyphens]{url}
  - \usepackage{microtype}
csl: apa.csl
--- 

The next part to narrow down a diferential expression analysis is to pre-process the data, which includes exploring the data, filtering, and normalizing it. In this section, we will explore the data and perform some pre-processing steps to prepare it for the differential expression analysis. In this part also is hellpfull to visuzalize the data to understand the conformation and state of the data, this with the principal porpose of identify any important variable that has to be included in the model to avoid confounding effects.87777

```{r}
#first of all we load the object and libraries that we required
library(recount3)
library(edgeR)
library(ggplot2)
library(SummarizedExperiment)
library(iSEE)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(variancePartition)

load("Results/objects_R/rse_gene_SRP119675_parsed.Rdata")
```

## Exploring the data
In this section we try to visualize the distribution of the data according to differents groups related with our qualitative variables, for this purpose we can use boxplots, density plots, and PCA plots to visualize the distribution of the data and identify any potential outliers or batch effects. This is related, for example with the plate the mise were processed, or view diference between groups with different phenotypes in our dataset, even in groups with the same phenotype but different sex or age. 

### Filtering 

#### Number of reads per group 
It is important to start filtering the counts that have a low proportion in pur analysis, this with the objective of do not consider those counts that has no stadistic power because it´s poor representation due to possible biased in the sequencing reaction, like bias related to the size of the reads. 

```{r}
#We calculate the proportion of the genes based on the total (normalizing from teh real scale to 0-1 scale)
rse_gene_SRP119675$assigned_gene_prop <- rse_gene_SRP119675$recount_qc.gene_fc_count_all.assigned /
  rse_gene_SRP119675$recount_qc.gene_fc_count_all.total

#Obtain the quartiles of our distribution
summary(rse_gene_SRP119675$assigned_gene_prop)
```

With this we can saw that the mean and the medium value are prety close and it does not seem to be a biased to one of the ways, but the objective is to eliminate those tha are poorly represented, and for this we can focus on those that are at the minium of proportion, for that we can make a histogram and dots plot for vizualiazting this mins values. 

```{r}
#We transfrom the colData information as a data frame to plot it using ggplot2
df <- as.data.frame(colData(rse_gene_SRP119675))

#Fisrt we plot a line with the gene proportions accounted by the data
ggplot(
  df,
  aes(x = reorder(rownames(df), assigned_gene_prop), y = assigned_gene_prop)
) +
  geom_point(alpha = 0.6) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(
    title = "Assigned Gene Proportion per Sample",
    x = "Samples (ordered)",
    y = "Assigned Gene Proportion"
  )


#We make the histogram of the same proportions to view in a better way
ggplot(df, aes(x = assigned_gene_prop)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "white",
    alpha = 0.8
  ) +
  labs(
    title = "Distribution of Assigned Gene Proportion",
    x = "Assigned Gene Proportion",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

We can observed that there are some values that desviate from the distribution, in this case are those that are bellow the value of the first Quartile with a treshold of (0.810), so we can filter that samples with low gene proportion coverage. 

```{r}
#show that genes that are under the treshold
cat(
  "Muestras que se encuentran por debajo del humbral de 0.810: ",
  table(rse_gene_SRP119675$assigned_gene_prop < 0.810)[2]
)

rse_gene_SRP119675_filtered <- rse_gene_SRP119675[,
  rse_gene_SRP119675$assigned_gene_prop > 0.810
]
```

#### Expression levels  
Also is important to not consider those counts that do not apport stadistical power to the DE analysis for this we can achive this with an stadistical backgruond provided by the fucntion `edgeR::filterByExpr()`, this is important to account with thse genes that are relevant in our groups that we selected important for modeling the diference in the expresion.

```{r}
#we filter the levels of expression based on the age groups
keep <- filterByExpr(
  assay(rse_gene_SRP119675_filtered),
  group = colData(rse_gene_SRP119675_filtered)$sra_attribute.genotype
)
rse_gene_SRP119675_filtered <- rse_gene_SRP119675_filtered[keep, ]

```


### Normalization 
For the final step is important to Normalize the data this in order to have equivalence in expression levels acroos the data and elimanting the *composition bias*, for this issue we are using the `DGEList` function of the edgeR library 


```{r}
dge <- DGEList(
  counts = assay(rse_gene_SRP119675_filtered, "counts"),
  genes = rowData(rse_gene_SRP119675_filtered)
)
dge <- calcNormFactors(dge)

logCPM <- edgeR::cpm(dge, log = TRUE)
assay(rse_gene_SRP119675_filtered, "logCPM") <- logCPM

```


### iSEE 
One good alternative to explore the data in an interactive way is to use the library `ìSEE` that use the obects of summarize experiment to make panels of diferent types of plots

Because we do not have all the time the reduction of dimentioanlity analysis we can do that converting our object to a SingleCellExperiment to perform the following reduction of dimentionality analysis 


```{r}
#for ploting with isee
sce_gene_SRP119675_filtered <- as(
  rse_gene_SRP119675_filtered,
  "SingleCellExperiment"
)
```

#### PCA
```{r}
#we have to transfrom our sumarized experiment object to a single cell experiment object to compute the principal components
#now we run de PCA
sce_gene_SRP119675_filtered <- scater::runPCA(
  sce_gene_SRP119675_filtered,
  exprs_values = "logCPM",
  ncomponents = 10
)

se <- sce_gene_SRP119675_filtered
colormap <- ExperimentColorMap()
se <- iSEE::cleanDataset(se)
colormap <- synchronizeAssays(colormap, se)

#plot the PCA
red.dim <- reducedDim(se, "PCA")
plot.data <- data.frame(
  X = red.dim[, 1],
  Y = red.dim[, 2],
  row.names = colnames(se)
)

plot.data$ColorBy <- colData(se)[, "sra_attribute.genotype"]

# Avoid visual biases from default ordering by shuffling the points
set.seed(142)
plot.data <- plot.data[sample(nrow(plot.data)), , drop = FALSE]

ggplot() +
  geom_point(
    aes(x = X, y = Y, color = ColorBy),
    alpha = 1,
    plot.data,
    size = 1
  ) +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "sra_attribute.genotype",
    title = "PCA"
  ) +
  coord_cartesian(
    xlim = range(plot.data$X, na.rm = TRUE),
    ylim = range(plot.data$Y, na.rm = TRUE),
    expand = TRUE
  ) +
  scale_color_manual(
    values = colDataColorMap(
      colormap,
      "sra_attribute.genotype",
      discrete = TRUE
    )(4),
    na.value = 'grey50',
    drop = FALSE
  ) +
  scale_fill_manual(
    values = colDataColorMap(
      colormap,
      "sra_attribute.genotype",
      discrete = TRUE
    )(4),
    na.value = 'grey50',
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = 'bottom',
    legend.box = 'vertical',
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    title = element_text(size = 12)
  )

```


#### MDS

```{r}
mat <- edgeR::cpm(dge, log = TRUE)
dist_mat <- dist(t(mat))
mds <- cmdscale(dist_mat, k = 2)

reducedDims(sce_gene_SRP119675_filtered)$MDS <- mds

red.dim <- reducedDim(sce_gene_SRP119675_filtered, "MDS")
plot.data <- data.frame(
  X = red.dim[, 1],
  Y = red.dim[, 2],
  row.names = colnames(se)
)

plot.data$ColorBy <- colData(se)[, "sra_attribute.genotype"]

# Avoid visual biases from default ordering by shuffling the points
set.seed(142)
plot.data <- plot.data[sample(nrow(plot.data)), , drop = FALSE]

ggplot() +
  geom_point(
    aes(x = X, y = Y, color = ColorBy),
    alpha = 1,
    plot.data,
    size = 1
  ) +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "sra_attribute.genotype",
    title = "MDS"
  ) +
  coord_cartesian(
    xlim = range(plot.data$X, na.rm = TRUE),
    ylim = range(plot.data$Y, na.rm = TRUE),
    expand = TRUE
  ) +
  scale_color_manual(
    values = colDataColorMap(
      colormap,
      "sra_attribute.genotype",
      discrete = TRUE
    )(4),
    na.value = 'grey50',
    drop = FALSE
  ) +
  scale_fill_manual(
    values = colDataColorMap(
      colormap,
      "sra_attribute.genotype",
      discrete = TRUE
    )(4),
    na.value = 'grey50',
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = 'bottom',
    legend.box = 'vertical',
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    title = element_text(size = 12)
  )

```

### Variance Partition 

As we can note neither the PCA nor the MDS could explain the variance in the expression levels previous the DE analysis so we can do a secondary analisis of `variancePartition` to acces the task of identifing the more relevant variables in our data, in other words those variables that explain in high level the variance in our data. 

#### Variance explicated for each variable  

```{r}
form <- form <- ~ sra_attribute.genotype +
  sra_attribute.age +
  sra_attribute.gender
varPart <- variancePartition::fitExtractVarPartModel(
  logCPM,
  form,
  colData(rse_gene_SRP119675_filtered)
)
vp <- sortCols(varPart)

plotVarPart(vp)
```

We can see that the most variance in our dataset is explained by the individual variance acroos the samples followed by the variance of the age and genotype, something important to see of these plots is that some genes variance is explained by the age as well as the genotype, that genes could no represent the total variance across samples an that was beacause we could not separate the clusters in the dimentional reduction analysis like PCA or MDS. 

#### Correlated variables 

```{r}
form <- ~ (1 | sra_attribute.genotype) +
  (1 | sra_attribute.age) +
  (1 | sra_attribute.gender) +
  (1 | sra_attribute.source_name)

C <- canCorPairs(form, colData(rse_gene_SRP119675_filtered))

plotCorrMatrix(C)

#for variance explication of a particular gene plotPercentBars(vp[gene, ])
```

We can see that the variable `sra_attribute.source_name` is highly correlated with the other varibles so we can drop out from de DE analysis

### Boxplots 

Finaly we can see the boxplot distribution od the samples, this can set a milestone of the aparently diference in the gene distribution acroos the genotypes in this case 

```{r}
ggplot(
  as.data.frame(colData(rse_gene_SRP119675_filtered)),
  aes(
    y = assigned_gene_prop,
    x = sra_attribute.genotype,
    fill = sra_attribute.genotype
  )
) +
  geom_boxplot() +
  theme_bw(base_size = 20) +
  ylab("Assigned Gene Prop") +
  xlab("Age Group") +
  scale_fill_discrete(name = "Genotype")

```

With this data preprocces coupled with the varianace analyses of the genes acroos the available variables we can have an idea of what we should consider in our mode, In this case we can consider the genotype, age and gender those that explain the more variance and do not are correlated with each other. 
